Lyman-$\alpha$ photons through rotating outflows

### 2018-04-02 paper/paper.tex
british package to break words in line ends
minor changes

\documentclass[a4paper,fleqn,usenatbib]{mnras}
%=========================================================================
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{grffile}
\usepackage{float}
\usepackage[dvips]{epsfig}
\usepackage{epsfig}
\usepackage{dblfloatfix}
\usepackage{color}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{bm}
%Non reposionated tables
\newcommand{\HI}{{\text{H\MakeUppercase{\romannumeral 1}}} }
\newcommand{\lya}{\ifmmode{{\rm Ly}\alpha}\else Ly$\alpha$\ \fi}
\newcommand{\kms}{\ifmmode\mathrm{km\ s}^{-1}\else km s$^{-1}$\fi}
\newcommand{\vrot}{\ifmmode v_{\mathrm{rot}}\else $v_{\mathrm{rot}}$~\fi}
\newcommand{\vout}{\ifmmode v_{\mathrm{out}}\else $v_{\mathrm{out}}$~\fi}
\newcommand{\tauh}{\ifmmode \tau_{\mathrm{H}}\else $\tau_{\mathrm{H}}$~\fi}
\newcommand{\vth}{\ifmmode v_{\mathrm{th}}\else $v_{\mathrm{th}}$~\fi}

@@

What is then the expected imprint of rotation on a resonant emission
line such as the \lya line? To what extent is possible to constrain
rotational kinematics from the \lya emission line? 
Detailed radiative transfer (RT) \lya modeling of rotating systems
started until recently by \cite{Garavito14}.
In that work the authors studied the influence of pure solid body
rotation on the \lya line's morphology.
They found that rotation indeed introduces changes, the most
noticeable being the dependence of the spectra with the viewing
angle with respect to the rotation axis.

@@

However, the gas dynamics in Lyman Alpha Emitter (LAE) galaxies is
more complex than pure rotation.
The case for outflows as traced in the \lya line is much better
established.
In many observations the \lya line profiles has a single peak
redwards from the line's center, in other cases there is a double peak
but the peak on the red side is stronger
\citep[e.g.][]{2010ApJ...717..289S,Erb14,Trainor16}.   
These features been readily explained as the consequence of multiple
\lya photon scatterings through an homogeneous outflowing shell of
neutral Hydrogen
\citep{2006A&A...460..397V,Orsi12,2012ApJ...751...29Y,2015ApJ...812..123G}.

An study of the combined effects of outflows and rotation has not been
presented in the literature.
Here we present such a study. 
We investigate a simplified geometrical configuration corresponding to a
spherical gas cloud with symmetrical radial outflows and solid body
rotation. 
We base our modeling on a Monte-Carlo radiative transfer code.
Besides modeling the impact of joint rotation and outflows, we also
check to what extent the analytical model presented by
\cite{Garavito14} to explain the effects of rotation can also be
applied in our case.

@@

\begin{figure}
\centering
\includegraphics[width=0.44\textwidth]{muse_vs_simulations.pdf}

@@

The Monte Carlo code we use (CLARA) \citep{CLARA} follows the propagation of
individual photons through a neutral Hydrogen medium characterized by
its temperature, velocity field and global optical depth.
The code assumes an homogeneous density throughout the simulated
volume.
In the current implementation we neglect the influence of dust.
Our basic model is an spherical distribution of neutral hydrogen,
an approximation commonly used in the literature, as it explains a
wide variety of observational features
\citep{Ahn03,Verhamme06,Dijkstra06}.

The velocity field we use captures both outflows and rotation.
Outflows are described by a Hubble-like radial velocity profile with
the velocity magnitude increasing linearly with the radial
coordinate; the outflows model is fully characterized by $v_{\rm out}$, the
velocity at the sphere's surface.
Rotation follows a solid body rotation profile, which is fully
characterized by $V_{\rm rot}$, the linear velocity at the sphere's surface.

@@

\begin{equation}
	v_{x}=\frac{x}{R}v_{\rm out}-\frac{y}{R}v_{\rm rot} ,
	\label{eq:vx}
\end{equation}

\begin{equation}
	v_{y}=\frac{y}{R}v_{\rm out}+\frac{x}{R}v_{\rm rot} ,
	\label{eq:vy}
\end{equation}

\begin{equation}
	v_{z}=\frac{z}{R}v_{\rm out},
	\label{eq:vz}
\end{equation}
%
where $x$, $y$ and $z$ are the cartesian position coordinates with the
origin at the sphere's center, $R$ is the radius of the sphere and the
direction of the angular velocity vector corresponds to the $\hat{k}$
unit vector.

For each model setup we follow $10^5$ individual photons generated at
the center of the sphere at the \lya\ line's center as they propagate
through the volume and finally scape.
We store the final frequency and propagation direction for each photon
at its last scattering.

As input parameters we use \tauh$=\{10^5, 10^6, 10^7\}$,
\vout$=\{5,25,50\}\kms$ and \vrot$=\{0,50,100\}\kms$, for a total of
$27$ models with all the possible parameter combinations.
The range of values have some overlap with the expectations from a
galaxy with a total neutral hydrogen mass of $10^8$-$10^9$
$M_{\odot}$.

\cite{Garavito14} presented an analytical model that
accounts for the effects of pure rotation on the
\lya\ line morphology. 
The basic assumption of their analytical model is that each
differential surface element on the sphere Doppler shifts the photons
that it emits.
In this paper we introduce this ansatz by post-processing the results
of the outflows simulations without rotation.
The frequency of each photon is Doppler shifted as follows

\begin{equation}
x' = x + \frac{\vec{v}_{\rm rot}\cdot\hat{k}}{v_{\rm th}}
\label{eq:shift_x}
\end{equation}
%
where $x'$ is the photon's new adimensional frequency, $x$ is the photon's
frequency after being processed by the outflow, $v_{\rm rot}$ is the 
rotational velocity at the point of escape of the photon, $\hat{k}$ is
the photon's direction of propagation and $v_{\rm th}$ is the thermal
velocity of the sphere.

@@

\section{Results}

\label{sec:results}

\subsection{Qualitative Trends}
\label{sec:qualitative}
Figure \ref{fig:doppler_shift} summarizes the most important trends.
The six panels correspond to $\tau=10^6$ and a viewing angle of
$\theta =90^{\circ}$, that is, perpendicular to the rotation axis of the
galaxy. 
In every panel the thin black line corresponds to the pure outflow
solution, i.e. without rotation. 
From top to bottom we see the effect of increasing the outflow
velocity, which is the expected increasing asymmetry towards the red
peak.

@@

The thick gray line corresponds to the pure outflow solution
with the Doppler boost added to model rotation's influence.
At $\tauh=10^6$ the Doppler boost does a good job at capturing the broad
morphological features introduced by rotation: the angle dependence,
the broadening and the intensity increase at the line's center.

In Figure \ref{fig:doppler_shift_theta_0} we show the same results as
in Figure \ref{fig:doppler_shift} but for a viewing angle of $\theta =
90^{\circ}$, that is parallel to the rotation axis. 
In this case we confirmethe result presented by \cite{Garavito14},
namely that pure rotation introduces a strong dependence with 
viewing angle, a trend that we find also holds for rotation mixed with
ouflows.

The quality of the results from the Doppler boost improves for higher
\tauh values. 
In the Appendix we show the same plots as Figures
\ref{fig:doppler_shift} and \ref{fig:doppler_shift}, there it is
evident that for $\tauh=10^5$ the results are not as good as they are
for $\tauh=10^6$, and that for $\tauh=10^7$ the Doppler boos
provides a remarkable good approximation.

@@

After finding the quantitative influence of the different parameters we
move onto a qualitative study.
To do this we summarize the line morphology by four different
characteristics: standard deviation (STD), skewness (SKW), bimodality
(BI) and valley/peak ration.
These quantities are defined by the following equations \citep{kokoska1999}.:

\begin{equation}
\label{eq:std}
\mathrm{STD} = \sqrt{m_2},
\end{equation}

\begin{equation}
\label{eq:skw}
\mathrm{SKW} = \frac{m_3}{m_2^{3/2}},
\end{equation}

\begin{equation}
\label{eq:bi}
\mathrm{BI} = \mathrm{KURTOSIS} - \mathrm{SKW}^2 = \frac{m_4}{m_2^{2}} - \frac{m_3^2}{m_2^{3}},
\end{equation}
%
where each $m_i$ is the i-th moment about the mean. 
The STD has velocity units and quantifies the line's width.
The SKW is adimensional and quantifies the peaks' asymmetry. 
In the case of a bimodal distribution, $SKW>0$ means that the blue
peak is taller and for $SWK<0$ the red peak is taller. 
The BI is adimensional and quantifies whether the line has 1 or 2
peaks: it is  always $\geq 1$ \citep{Pearson1929} and the 
to 1, the more bimodal is the line (i.e. has 2 similar peaks). 
We found by visual inspection of our spectra that BI$=2.5$ marks the
transition between two peaks (however imbalanced) and a dominant
single peak.

@@

Figure \ref{fig:bimodality} shows the results for the bimodality using
the same layout as in the two previous Figures.
Following the reasoning about the skewness, we observe that
increasing the outflow velocity increases the value of bimodality,
that is, it transitions to a more pronounced single peak. 
The trend as a function of the rotational velocity and the optical
depth are not monotonous.
When the outflow velocity is low (\vout$<50\ \kms$), an increasing
rotational velocity smears the two asymmetrical peaks pushing the line
morphology towards a single peaks, making the bimodality statistics
increase. 
On other situations (\vout$=50\ \kms$ and \tauh$\geq 10^6$) higher
rotational velocities the bimodality statistics decreases, which means
that it manages to slightly enlarge the already dominant red peak.

@@

In Figure \ref{fig:valley_intensity} we quantify how the intensity at
the line's center (i.e. the valley) changes with the viewing angle,
the outflow velocity and the optical depth.
These results correspond to a fixed rotational velocity of
\vrot$=100\ \kms$.
The triangles correspond to the RT simulations and the line represents
the DS results. 
The valley intensity is expressed as a fraction of the maximum peak
intensity in the line, as such the valley/peak ratio is always less
than one. 
In every panel we see that the valley/peak ratio decreases as the
observer moves from a line  of sight perpendicular to the rotation
axis onto a parallel line of sight. 
This is a clear demonstration of the viewing angle dependency
introduced by rotation.

The valley/peak ratio at $\cos\theta=1$ matches results
without rotation, this shows that for increasing rotational velocity
the valley/peak ratio increases.
In turn, for increasing optical depth or outflowing velocity this
ratio decreases.
Once again, the DS results correctly follow the trends for the full RT
simulations.  
This time the differences have an upper bound of $55\%$, $2\%$ and
$1\%$ for  $\tauh=10^5$, $10^6$ and $10^7$, respectively.

@@

\subsubsection{\lya continuum leakage or rotation?}

Another application of this new model to the current interpretation of \lya
spectra, is that the central ($V=0$) emission of the spectra seen in the
\lya line, is a consequence of the viewing angle of the galaxy and can be
controlled by it. 
Several authors have suggested that this central emission is caused by
radiation that escapes the galaxy without scattering, mostly due to a
clumpy interstellar medium \citep{Hansen06, 2016ApJ...833L..26G}.
Rotation is an alternative that solves this issue.

We note that the relative intensity of the two peaks in the outflowing
spectra is not modified by rotation, that asymmetry is solely
controlled by the outflow velocity.
This means that results that have already derive a typical outflowing
velocity as to match the observational constraints do not have to be
completely revised. 
Only an additional exploration of rotational velocities and viewing
angles need to be explored.

A recent observational example is provided by
\cite{2017A&A...608L...4R}.
They report optical spectroscopy of a bright lensed galaxy at a
redshift of $z=2.4$. 
The object presents a triple peaked \lya profile that can be thought
as the superposition of two components, a double asymmetric peak and a
narrower symmetric peak around the line's center.
One possible option for the narrow component is that it escaped the
galaxy without substantial interaction with neutral Hydrogen.
We speculate that the central feature could also be due to a purely 
rotating component that can produce such a feature as shown by
\cite{tololo}. The triple peak could be then produced by two main
decopled components: one rotating the other outflowing.

Another interesting example are the \lya lines in Green Pea galaxies \citep{2016ApJ...820..130Y}.
\cite{2017ApJ...844..171Y} showed in a study of 43 Green Pea galaxies
that 2/3 were strong \lya emitters. 
There the \lya escape fraction is effectively estimated from the ratio
of H$\alpha$ to \lya.
Galaxies with large escape fractions also substantial \lya emission at
the line's center. 
We speculate that this feature could help to constrain rotation in
Green Pea galaxies.

\subsubsection{\lya Kinematic Maps}

Current observational facilities have the capability of spatially
resolving the extent of a LAE.
For instance \cite{Prescott14} presented observational results of a
Doppler shift when taking spectra at two oposite sides of a large
($\approx 80$kpc) LAE.
In more recent work \cite{2018MNRAS.473.3907A} mapped \lya emission
around a quasar

@@

Although it is a good approximation to think the rotating spectra by a
sum of Doppler shifts, the peak of the spectra is also weighted by the
amount of mass with a given velocity. 
In this toy model the distance between the peaks of the
receding/approaching spectra is close to $\sim 20\ \kms$, which is
less than half of the naively expected value of $2\vrot=100\ \kms$,
due to the fact that only a small fraction of the photons are emitted
at the extreme of the galaxy having the maximum rotational velocity of
$50\ \kms$.

New spectrographs like MUSE should help obtain kinematic information
from LAEs to build velocity maps in \lya. 
This could be a natural extension of the work reported by
\cite{Herenz2016} on the velocity maps of several LARS (Lyman Alpha
Reference Sample) galaxies.  
The interpretation of such data should take into account the insights
and trends we have presented in this paper.

\subsubsection{Comparison against the MUSE-Wide Survey}

@@

In this paper we explore, for the first time in the literature,
the results of a model for the emergent \lya\ line from rotating outflows.
The main results for the model are computed from a Monte-Carlo
radiative transfer simulation and confronted with a simple
semi-analytic ansatz that adds the effects of rotation onto results of
pure outflow kinematics.

To address the first question we posed in the introduction (\emph{what is then
the expected imprint of rotation on a resonant emission line such as
the \lya line?}) we can say that the main effects of rotation on the
\lya\ line morphology are:

@@

We find that these effects can be quantitatively explained by a
Doppler shift computed as product of quantities at the surface of last
scattering, namely  $\vec{v}_{\rm rot}\cdot \hat{k}$,  where
$\vec{v}_{\rm rot}$ is the velocity due to rotation and $\hat{k}$ is
the direction of the photon's propagation.  
These conclusions, specially the confirmation of the Doppler Boost as
a good approximation to model solid body rotation, strengthen the
semi-analytic model reported by \cite{Garavito14} when they modeled
the effect of pure rotation on \lya\ spectra.

Addressing the second question we posed in the introduction (\emph{to what
extent is possible to constrain rotational kinematics from the \lya
emission line?})  we can say that as an application to observational data our results suggest that
rotation should also be considered as an explanation of the
non-vanishing \lya intensity at the line's center, which could change
the interpretation of recent observational results
\citep[e.g.][]{2017ApJ...844..171Y,2017A&A...608L...4R} that claim
instead that this flux constrains the escape fraction of \lya
radiation.

A more straightforward approach takes into account how the
\lya spectra from two different sides of a galaxy can detect
approaching/receding gas motions.
In that case one has to be careful and take into account that the
peaks of these two spectra is smaller than the expected value of
$2\times\vrot$ due to the different weights at the surface of the
emitting regions.

We also use data from MUSE to demonstrate to what extent the
outputs of the RT results presented here can be expected to match
hundreds of observed \lya spectra, but a full constraint of rotation
properties certainly require a full Markov Chain Monte Carlo
exploration of parameter space.

Our work confirms that rotation produces features in the \lya line
strong enough that can be expected to be detected in observations and
that should be taken into account into \lya modeling; the Doppler
Boost offers an easy-to-implement approximation to explore such
influence into already existing RT simulations and provide stronger
footing to the interpretation of \lya kinematics maps
\citep[e.g][]{2018MNRAS.473.3907A}.

>>>>

\documentclass[a4paper,fleqn,usenatbib]{mnras}
%=========================================================================
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{grffile}
\usepackage{float}
\usepackage[dvips]{epsfig}
\usepackage{epsfig}
\usepackage{dblfloatfix}
\usepackage{color}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{bm}
\usepackage[british]{babel}
%Non reposionated tables
\newcommand{\HI}{{\text{H\MakeUppercase{\romannumeral 1}}} }
\newcommand{\lya}{\ifmmode{{\rm Ly}\alpha}\else Ly$\alpha$\ \fi}
\newcommand{\kms}{\ifmmode\mathrm{km\ s}^{-1}\else km s$^{-1}$\fi}
\newcommand{\vrot}{\ifmmode v_{\mathrm{rot}}\else $v_{\mathrm{rot}}$~\fi}
\newcommand{\vout}{\ifmmode v_{\mathrm{out}}\else $v_{\mathrm{out}}$~\fi}
\newcommand{\tauh}{\ifmmode \tau_{\mathrm{H}}\else $\tau_{\mathrm{H}}$~\fi}
\newcommand{\vth}{\ifmmode v_{\mathrm{th}}\else $v_{\mathrm{th}}$~\fi}
\newcommand{\hatk}{\ifmmode \hat{k}\else $\hat{k}$~\fi}
\newcommand{\STD}{\ifmmode \mathrm{STD}\else $\mathrm{STD}$~\fi}
\newcommand{\SKW}{\ifmmode \mathrm{SKW}\else $\mathrm{SKW}$~\fi}
\newcommand{\BI}{\ifmmode \mathrm{BI}\else $\mathrm{BI}$~\fi}

@@

What is then the expected imprint of rotation on a resonant emission
line such as the \lya line? To what extent is it possible to constrain
rotational kinematics from the \lya emission line? 
Detailed radiative transfer (RT) \lya modeling of rotating systems
started until recently by \cite{Garavito14}.
In that work the authors studied the influence of pure solid body
rotation on the \lya line's morphology.
They found that rotation indeed introduces changes, the most
noticeable being the dependence of the spectra with the viewing
angle with respect to the rotation axis.

@@

However, the gas dynamics in Lyman Alpha Emitter (LAE) galaxies are
more complex than pure rotation.
The case for outflows as traced in the \lya line is solidly
established.  
In many observations the \lya line profile has a single peak
redwards from the line's center, in other cases there is a double peak
but the peak on the red side is stronger
\citep[e.g.][]{2010ApJ...717..289S,Erb14,Trainor16}.   
These features have been explained as the consequence of multiple
\lya photon scatterings through a homogeneous outflowing shell of
neutral Hydrogen
\citep{2006A&A...460..397V,Orsi12,2012ApJ...751...29Y,2015ApJ...812..123G}.

Nevertheless, an study of the combined effects of outflows and
rotation has not been presented in the literature.
Here we report on such a study. 
We investigate a simplified geometrical configuration corresponding to a
spherical gas cloud with symmetrical radial outflows and solid body
rotation. 
We base our modeling on a Monte-Carlo radiative transfer code.
Besides modeling the impact of joint rotation and outflows, we also
check to what extent the semi-analytical model presented by
\cite{Garavito14} an also be applied in our case.

@@

\begin{figure}
\centering
\includegraphics[width=0.45\textwidth]{muse_vs_simulations.pdf}

@@

The Monte Carlo code we use (CLARA) \citep{CLARA} follows the propagation of
individual photons through a neutral Hydrogen medium characterized by
its temperature, velocity field and global optical depth.
The code assumes a homogeneous density throughout the simulated
volume.
In the current implementation we neglect the influence of dust.
Our basic model is an spherical distribution of neutral hydrogen,
an approximation commonly used in the literature, as it explains a
wide variety of observational features
\citep{Ahn03,Verhamme06,Dijkstra06}.

The velocity field we use captures both outflows and rotation.
Outflows are described by a Hubble-like radial velocity profile with
the velocity magnitude increasing linearly with the radial
coordinate; the outflows model is fully characterized by $v_{\rm out}$, the
radial velocity at the sphere's surface.
Rotation follows a solid body rotation profile, which is fully
characterized by $v_{\rm rot}$, the linear velocity at the sphere's surface.

@@

\begin{equation}
	v_{x}=\frac{x}{R}\vout - \frac{y}{R}\vrot ,
	\label{eq:vx}
\end{equation}

\begin{equation}
	v_{y}=\frac{y}{R}\vout + \frac{x}{R}\vrot ,
	\label{eq:vy}
\end{equation}

\begin{equation}
	v_{z}=\frac{z}{R}\vout,
	\label{eq:vz}
\end{equation}
%
where $x$, $y$ and $z$ are the cartesian position coordinates with the
origin at the sphere's center, $R$ is the radius of the sphere and the
direction of the angular velocity vector is the $z$ axis.

For each model setup we follow $10^5$ individual photons generated at
the center of the sphere at the \lya line's center as they propagate
through the volume and finally scape.
We store the final frequency and propagation direction for each photon
at its last scattering.

As input parameters we use $\tauh=\{10^5, 10^6, 10^7\}$,
$\vout=\{5,25,50\}\;\kms$ and $\vrot=\{0,50,100\}\;\kms$, for a total of
$27$ models with all the possible parameter combinations.
The range of values have some overlap with the expectations from a
galaxy with a total neutral hydrogen mass of $10^8$-$10^9$
$M_{\odot}$.

\cite{Garavito14} presented an analytical model that
accounts for the effects of pure rotation on the
\lya line morphology. 
The basic assumption of their analytical model is that each
differential surface element on the sphere Doppler shifts the photons
that it emits.
In this paper we introduce this ansatz by post-processing the results
of the outflows simulations without rotation.
The frequency of each photon is Doppler shifted as follows

\begin{equation}
x' = x + \frac{\vec{v}_{\rm rot}\cdot \hatk}{\vth}
\label{eq:shift_x}
\end{equation}
%
where $x'$ is the photon's new adimensional frequency, $x$ is the photon's
frequency after being processed only by the outflow, \vrot is the 
rotational velocity at the point of escape of the photon, \hatk is
the photon's direction of propagation and \vth is the thermal
velocity of the sphere.
We also define the viewing angle, $\theta$, as the angle between the
rotation axis and the line of sight of a potential observer.

@@

\section{Results}
\label{sec:results}

\subsection{Qualitative Trends}
\label{sec:qualitative}

Figure \ref{fig:doppler_shift} summarizes the most important trends
from rhw RT simulations.
In the left side, the six panels correspond to $\tau=10^6$ and a viewing angle of
$\theta =90^{\circ}$, that is, perpendicular to the rotation axis of the
galaxy. 
In every panel the thin black line corresponds to the pure outflow
solution, i.e. without rotation. 
From top to bottom we see the effect of increasing the outflow
velocity, which is the expected increasing asymmetry towards the red
peak.

@@

The thick gray line corresponds to the pure outflow solution
with the Doppler shift added to model rotation's influence.
At $\tauh=10^6$ the Doppler shift does a good job at capturing the broad
morphological features introduced by rotation: the angle dependence,
the broadening and the intensity increase at the line's center.

In the right side of Figure \ref{fig:doppler_shift} we show the same results as in the left one, but for a viewing angle of $\theta =
0^{\circ}$, that is parallel to the rotation axis. 
In this case we confirm the result presented by \cite{Garavito14},
namely that pure rotation introduces a strong dependence with 
viewing angle, a trend that we find also holds for rotation mixed with
outflows.

The quality of the results from the Doppler shift improves for higher
\tauh values. 
In the Appendix we show the same plots as Figure \ref{fig:doppler_shift}, there it is
evident that for $\tauh=10^5$ the results are not as good as they are
for $\tauh=10^6$, and that for $\tauh=10^7$ the Doppler shift
provides a remarkable good approximation.

@@

After finding the qualitative influence of the different parameters we
move onto a quantitative study.
To do this we summarize the line morphology by four different
scalars: standard deviation (STD), skewness (SKW), bimodality
(BI) and valley/peak ratio.
These quantities are defined by the following equations \citep{kokoska1999}:

\begin{equation}
\label{eq:std}
\STD = \sqrt{m_2},
\end{equation}

\begin{equation}
\label{eq:skw}
\SKW = \frac{m_3}{m_2^{3/2}},
\end{equation}

\begin{equation}
\label{eq:bi}
\BI = \mathrm{KURTOSIS} - \SKW^2 = \frac{m_4}{m_2^{2}} - \frac{m_3^2}{m_2^{3}},
\end{equation}
%
where each $m_i$ is the i-th moment about the mean. 
The STD has velocity units and quantifies the line's width.
The SKW is adimensional and quantifies the peaks' asymmetry. 
In the case of a bimodal distribution, $\SKW>0$ means that the blue
peak is taller and for $\SKW<0$ the red peak is taller. 
The BI is adimensional and quantifies whether the line has 1 or 2
peaks: it is  always $\geq 1$ \citep{Pearson1929} and the closer 
to 1, the more bimodal is the line (i.e. has 2 similar peaks). 
We found by visual inspection of our spectra that $\BI=2.5$ marks the
transition between two peaks (however imbalanced) and a dominant
single peak.

@@

Figure \ref{fig:bimodality} shows the results for the bimodality using
the same layout as in the two previous Figures.
Following the reasoning about the skewness, we observe that
increasing the outflow velocity increases the value of bimodality,
that is, it transitions to a more pronounced single peak. 
The trend as a function of the rotational velocity and the optical
depth are not monotonous.
When the outflow velocity is low ($\vout<50\ \kms$), an increasing
rotational velocity smears the two asymmetrical peaks pushing the line
morphology towards a single peaks, making the bimodality statistics
increase. 
On other situations ($\vout=50\ \kms$ and $\tauh\geq 10^6$) higher
rotational velocities the bimodality statistics decreases, which means
that it manages to slightly enlarge the already dominant red peak.

@@

In Figure \ref{fig:valley_intensity} we quantify how the intensity at
the line's center (i.e. the valley) changes with the viewing angle,
the outflow velocity and the optical depth.
These results correspond to a fixed rotational velocity of
$\vrot=100\ \kms$.
The triangles correspond to the RT simulations and the line represents
the DS results. 
The valley intensity is expressed as a fraction of the maximum peak
intensity in the line, as such the valley/peak ratio is always $<1$. 
In every panel we see that the valley/peak ratio decreases as the
observer moves from a line  of sight perpendicular to the rotation
axis onto a parallel line of sight. 
This is a clear demonstration of the viewing angle dependency
introduced by rotation.

The valley/peak ratio at $\cos{\theta}=1$ matches results
without rotation, this shows that for increasing rotational velocity
the valley/peak ratio increases.
In turn, for increasing optical depth or outflowing velocity this
ratio decreases.
Once again, the DS results correctly follow the trends for the full RT
simulations.  
This time the differences have an upper bound of $55\%$, $2\%$ and
$1\%$ for  $\tauh=10^5$, $10^6$ and $10^7$, respectively.

@@

\subsection{\lya continuum leakage or rotation?}

One application of this new model to the current interpretation of \lya
spectra, is that the central ($V=0$) emission of the spectra seen in the
\lya line, is a consequence of the viewing angle of the galaxy and can be
controlled by it. 
Several authors have suggested that this central emission is caused by
radiation that escapes the galaxy without scattering, mostly due to a
clumpy interstellar medium \citep{Hansen06, 2016ApJ...833L..26G}.
Rotation is an alternative that solves this issue.

We note that the relative intensity of the two peaks in the outflowing
spectra is not modified by rotation, that asymmetry is solely
controlled by the outflow velocity.
This means that results that have already derive a typical outflowing
velocity as to match the observational constraints do not have to be
completely revised. 
Only an additional exploration of rotational velocities and viewing
angles needs to be explored.

A recent observational example is provided by
\cite{2017A&A...608L...4R}.
They report optical spectroscopy of a bright lensed galaxy at a
redshift of $z=2.4$. 
The object presents a triple peaked \lya profile that can be thought
as the superposition of two components, a double asymmetric peak and a
narrower symmetric peak around the line's center.
One possible option for the narrow component is that it escaped the
galaxy without substantial interaction with neutral Hydrogen.
We speculate that the central feature could also be due to a purely 
rotating component that can produce such a feature as shown by
\cite{tololo}. The triple peak could be then produced by two main
decoupled components: one rotating the other outflowing.

Another interesting example are the \lya lines in Green Pea galaxies \citep{2016ApJ...820..130Y}.
\cite{2017ApJ...844..171Y} showed in a study of 43 Green Pea galaxies
that 2/3 were strong \lya emitters. 
There the \lya escape fraction is effectively estimated from the ratio
of H$\alpha$ to \lya.
Galaxies with large escape fractions also substantial \lya emission at
the line's center. 
We speculate that this feature could help to constrain rotation in
Green Pea galaxies and find new correlations between \lya escape
fraction and galaxy rotation.

\subsection{\lya Kinematic Maps}

Current observational facilities have the capability of spatially
resolving the extent of a LAE.
For instance \cite{Prescott14} presented observational results of a
Doppler shift when taking spectra at two opposite sides of a large
($\approx 80$kpc) LAE.
In more recent work \cite{2018MNRAS.473.3907A} mapped \lya emission
around a quasar.

@@

Although it is a good approximation to think the rotating spectra by a
sum of Doppler shifts, the peak of the spectra is also weighted by the
amount of mass with a given line-of-sight velocity. 
In this toy model the distance between the peaks of the
receding/approaching spectra is close to $\sim 20\ \kms$, which is
less than half of the naively expected value of $2\vrot=100\ \kms$,
due to the fact that only a small fraction of the photons are emitted
at the extreme of the galaxy having the maximum rotational velocity of
$50\ \kms$.

Spectrographs like the Multi Unit Spectroscopic Explorer (MUSE) could
obtain kinematic information from large samples of LAEs to build
velocity maps in Ly$\alpha$.  
This could be a natural extension of the work reported by
\cite{Herenz2016} on the velocity maps of several LARS (Lyman Alpha
Reference Sample) galaxies.  
The interpretation of such data should take into account the insights
and trends we have presented in this paper.

\subsection{Comparison against the MUSE-Wide Survey}

@@

In this paper we explore, for the first time in the literature,
the results of a model for the emergent \lya line from rotating outflows.
The main results for the model are computed from a Monte-Carlo
radiative transfer simulation and confronted with a simple
semi-analytic ansatz that adds the effects of rotation onto results of
pure outflow kinematics.

To address the first question we posed in the introduction (\emph{what is then
the expected imprint of rotation on a resonant emission line such as
the \lya line?}) we can say that the main effects of rotation on the
\lya line morphology are:

@@

We find that these effects can be quantitatively explained by a
Doppler shift computed as product of quantities at the surface of last
scattering, namely  $\vec{v}_{\rm rot}\cdot \hatk$, where
$\vec{v}_{\rm rot}$ is the velocity due to rotation and \hatk is
the direction of the photon's propagation.  
These conclusions, specially the confirmation of the Doppler shift  as
a good approximation to model solid body rotation, strengthen the
semi-analytic model reported by \cite{Garavito14} where they modeled
the effect of pure rotation on \lya spectra.

Addressing the second question we posed in the introduction (\emph{to what
extent is it possible to constrain rotational kinematics from the \lya
emission line?})  we can say that as an application to observational data our results suggest that
rotation should also be considered as an explanation of the
non-vanishing \lya intensity at the line's center, which could change
the interpretation of recent observational results
\citep[e.g.][]{2017ApJ...844..171Y,2017A&A...608L...4R} and provide a
test for rotation in the observed systems.

A more straightforward approach takes into account how the
\lya spectra from two different sides of a galaxy can detect
approaching/receding gas motions \citep{Prescott14,2018MNRAS.473.3907A}.
In that case, one has to be careful and take into account that the
peaks of these two spectra is smaller than the expected value of
$2\times\vrot$ due to the different weights at the surface of the
emitting regions.

We also use data from MUSE \citep{2017A&A...606A..12H} to show
to what extent the outputs of the RT results presented here can be
expected to match hundreds of observed \lya spectra, but a full
constraint of rotational properties certainly requires a full Markov
Chain Monte Carlo exploration of parameter space.

Our work confirms that rotation produces features in the \lya line
strong enough that can be expected to be detected in observations and
that should be taken into account into \lya modeling; the Doppler
Shift offers an easy-to-implement approximation to explore such
influence into already existing RT simulations and provide stronger
footing to the interpretation of present and future \lya kinematics
maps \citep[e.g][]{2018MNRAS.473.3907A}.
